### 2012-08-01 Doc/VRIPHYS2012/vriphys2012.tex
sections 3, 4 and most of 5
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@443 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

% Partial derivative
\newcommand{\deriv}[2]{\frac{\partial #1}{\partial #2}}

@@

\section{Element Kinematics} % {{{

@@

\begin{equation}\label{bez-delta}
    \begin{split}
        \delta T = &   (n_1 + n_4 + n_5 + \frac{1}{3} n_{10})  \delta \mathbf{x}_1 + \\
                   & + (n_2 + n_6 + n_7 + \frac{1}{3} n_{10})  \delta \mathbf{x}_2 + \\
                   & + (n_3 + n_8 + n_9 + \frac{1}{3} n_{10})  \delta \mathbf{x}_3 + \\
                   & + \sum_{j=4}^9 n_j P_{i-j} \times \delta \mathbf{\theta}_i + \\
                   & + \frac{n_{10}}{3} \sum_{i=1}^3 P_{i-10} \times \delta \mathbf{\theta}_i 
    \end{split}
\end{equation}

@@

The term $\deriv{T^k}{\xi_i}$ can be derived from the expression of $T(\xi_1,\xi_2, \xi_3)$ in equation (\ref{global-transform}) and 
$\deriv{\xi_i}{l}$ from equation (\ref{shape-vector}). 
The in-plane displacement gradient can be obtained with tensor $F$:
%
\begin{equation}
\deriv{u^k}{l} = F_{kl} - I_{kl} 
\end{equation}
%
where $I$ is the identity matrix ($I_{kl}  = \{1$ if $l=k$ ;  $0$ if $l \neq k \} $). 
The displacement gradient can be linked to the degrees of freedom of the shell using the expression of the displacement of a point $\delta T$ (see equation  (\ref{bez-delta})) on the shell surface:
\begin{equation}
    \nabla u = \nabla(\delta T)
\end{equation}

@@

In order to compute FEM-based stiffness for the shell element, we need to compute the jacobian matrix between strains and nodes' displacements.
As we have separated the formulations for the in-plane and the out of plane displacements, it will result with the computation of two jacobian matrices (of size $3\times9$).
\begin{equation}
\delta \left[ \begin{array}[c]{c} \epsilon_{xx} \\ \epsilon_{yy}  \\ 2\epsilon_{yx} \end{array} \right] = 
\left[ \mathbf{J}_m \right] 
\left[ \begin{array}[c]{c} \delta \mathbf{x}_{1x} \\ \delta \mathbf{x}_{1y} \\ \delta \theta_{1z} \\ \vdots    \end{array} \right]
\end{equation}
where $\left[ \mathbf{J}_m \right]$ maps the membrane in-plane displacements and 
%
\begin{equation}
\delta \left[ \begin{array}[c]{c} \epsilon_{xx} \\ \epsilon_{yy}  \\ 2\epsilon_{yx} \end{array} \right] = 
\left[ \mathbf{J}_b \right] 
\left[ \begin{array}[c]{c} \delta \mathbf{x}_{1z} \\ \delta \theta_{1x} \\ \delta \theta_{1y} \\ \vdots    \end{array} \right]
\end{equation}
where $\left[ \mathbf{J}_b \right]$ maps the bending out-plane displacements.

To compute the values in matrices $\left[ \mathbf{J}_m \right]$ and $\left[ \mathbf{J}_b \right]$,  we can reuse equation (\ref{bez-delta}) and apply the spacial derivatives.
To illustrate, let's take one value of the matrix $\left[ \mathbf{J}_m \right]$  (first line, fourth column), we first take the kinematic between $\delta T_x$ and the concerned degree of freedom $ \delta \mathbf{x}_{2x}$ from equation (\ref{bez-delta}):
$$
   \frac{ \delta T_x}{ \delta \mathbf{x}_{2x}} = (n_2 + n_6 + n_7 + \frac{1}{3} n_{10})
$$
Then, we apply the spacial derivatives to compute the variation of $\epsilon_{xx}$ due to a variation of $\mathbf{x}_{2x}$.
$$
\delta \epsilon_{xx} += 
\underbrace{\left( \sum_{i=1}^{3} (\frac{\partial n_2}{\xi_i} + \frac{\partial n_6}{\xi_i} + \frac{\partial n_7}{\xi_i} + \frac{1}{3}\frac{\partial n_{10}}{\xi_i})\frac{\partial \xi_i}{\partial x} \right) }_{\left[ \mathbf{J}_m \right](1,4)}  
{\delta \mathbf{x}_{2x}}
$$

@@

\begin{align}
    \label{eq-Km}
    \mathbf{K}_m & = \iiint_V \mathbf{J}_m^T M \mathbf{J}_m \, \mathrm{d} V \\
    \label{eq-Kb}
    \mathbf{K}_b & = \iiint_V \mathbf{J}_b^T M \mathbf{J}_b \, \mathrm{d} V
\end{align}
%% comment
\CD{This is something that is maybe wrong in our formulation... as the bending adds a contribution on the deformation tensor, maybe we should not separate the contribution from membrane and bending as we do now ($K_m$ and $K_b$ are computed separately) }
%% comment
\CD{here, it would be... $\mathbf{K}_e =  \iiint_V \left[\begin{matrix} \mathbf{J}_m & \mathbf{J}_b \end{matrix} \right]^T  M \left[\begin{matrix} \mathbf{J}_m & \mathbf{J}_b \end{matrix} \right]$. Could we test that option easily in the code, or is it a nightmare ??}
%%

\noindent
where $M$ is the material matrix. To keep the system simple we use linear
Hooke's law for isotropic materials.

Because the deflection field for the element is non-linear \eqref{eq-delta}
in position the integrals \eqref{eq-Km} and \eqref{eq-Kb} have to be computed using
numerical integration. In our implementation we employ 6-point Gaussian
quadrature for integration over triangle area \CD{TODO: figure + gives the formula of the quadrature...}
% TODO: add ref
% TODO: add abscissas/weights?

@@

It is known that the Cauchy's strain tensor \eqref{eq-cauchy} is not
rotation invariant \footnote{add ref} and produces ghost forces for rigid
rotations. The alternative approach is to use Green's strain tensor,
however it leads to nonlinear system. To keep the system linear and to deal
with rigid body rotations we compute the displacements and forces in
corotational frame. At each step of the simulation we compute the local
frame for every element, the resulting internal force of deformed element
is then computed relative to this frame:

\begin{equation}
    F_e = R_e^T (K_e u) = R_e^T \left(K_e(R_e x - R_e^0 x^0)\right)
\end{equation}

\noindent
where $R$ and $R_0$ are rotation matrices of the frame for current and rest
mesh respectively.

TODO: how we compute frame?
%from the position of the corner nodes and we align one axis of the frame with one
%edge of the element. \footnote{add image}

%The displacement field defined in \eqref{eq-delta} is clearly symmetric in
%respect to deformations, but ...

This choice of the frame aligns one axis of the frame with one of the edges
of the triangle. Correct choice of the corotational frame is a tricky
problem \cite{Felippa2005} and incorrect frame for the membrane element can
have adverse effects on the simulation. In our case we have experienced
nonnegligible difference in deformation of equilateral triangle depending
on whether an axis was aligned to the edge favored by the deformation or to
one of the two other edges.

Performing a polar decomposition on the deformation gradient $F = \nabla u
+ I$ can be used to extract a rotation component from the deformation of
the element \cite{Ciarlet1994}. Because we are only fixing the membrane
element we can perform the polar decomposition in 2D with the following
formula:

\begin{equation}
    R = F + \mathrm{sgn}(\mathrm{det}(F)) \begin{pmatrix}
        F_{22} & - F_{21} \\
        F_{12} & F_{11}
    \end{pmatrix}
\end{equation}

\noindent
and subsequently normalizing the columns. We then rotate the corotational
frame in inverse direction. Even single step of this process improves the
final frame but more iterations of the polar decompositions are usually
necessary (up to 20) which severely degrades the performance. But polar
decomposition tends to "overshoot" the ideal rotation angle and each
subsequent step performs a rotation in the opposite direction. By properly
scaling the rotation angle we are able to minimize the amount of needed
iterations down to 5. The best value for the scaling factor depends on the
simulated problem, but we have experimentally localized the best value to
be somewhere between 0.6 and 0.65. We choose to use the value 0.61.

We employ two stop conditions for the iteration process:
\begin{itemize}
    \item The change in rotation angle is less than $10^{-6}$,
    \item and the maximum number of 5 iterations.
\end{itemize}

If the first condition is not met in 5 iterations it can mean one of three
things:

\begin{enumerate}
    \item We still have very good approximation of the ideal frame.
    \item The convergence for the element is too slow.
    \item The process is diverging.
\end{enumerate}

\noindent
and performing any more iterations is unnecessary.

@@

\section{Visual and Mechanical Mapping} % {{{

Because of the bending property of the shells relatively few elements are
necessary to simulate curved surface. To enrich the visual experience from
the simulated object it is desirable to use more triangles in the areas
with high curvature during rendering. Besides the visual accuracy the high
resolution mesh correctly modeling the curved areas is needed for
interaction and correct simulation of collisions and constraints. To handle
these we have also to know how to map the velocities from the mechanical
mesh onto the high resolution mesh and how to map forces acting on the high
resolution mesh back onto the mechanical mesh.

\subsection{Mapping the Surface Mesh}

We use the fact that the geometry of the shell is based on the formulation
of cubic B\'ezier triangle \eqref{eq-cubicbez}. For every vertex of the high
resolution mesh we first find the corresponding triangle on the mechanical
mesh that is closest to the vertex and assign barycentric coordinates on
the triangle to this vertex. After every step of the simulation the high
resolution mesh is updated using the assigned barycentric coordinates and
the function of the surface \eqref{eq-cubicbez}.

\subsection{Mapping Velocities}

In the

Differentiating \eqref{eq-cubicbez2} by time we get:

\begin{equation}
    \dot{T} = \sum_{i=1}^{10} n_i \dot{P_i} = \sum_{i=1}^{10} n_i V_i
\end{equation}

Which means we use the same expression to interpolate the velocities only
substituting the velocities at control points. For the corner nodes we
already have the velocities $V_i$. For the internal nodes we need to compute
them. In case of the edge nodes it is a problem of computing the velocity
of a point attached to the moving rigid body:

\begin{eqnarray}
    V_j = V_i + \omega_i \times (P_j - P_i)
\end{eqnarray}

where $j$ is the index of the corner node this edge node is attached to and
$\omega_j$ is the angular velocity at the node. For the central node we
the formula is again similar to \eqref{eq-central2}:

\begin{eqnarray}
    V_i = \frac{1}{3} \sum_{i=1}^3 V_i + \omega_i \times (P_{10} - P_i)
\end{eqnarray}

\subsection{Mapping Forces}

For physical interaction with the object one also requiers that the normal
forces applied on the high resolution mesh are trasmitted back as forces
and torques on the vertices of the mechanical mesh.

We compute the influence of force $F$ on the corners through all the
associated control points by the formula:

\begin{eqnarray}
    F_1 & = & F \sum_{i \in \{1,4,5\}} n_i + \frac{1}{3} F n_{10} \\
    F_2 & = & F \sum_{i \in \{2,6,7\}} n_i + \frac{1}{3} F n_{10} \\
    F_3 & = & F \sum_{i \in \{3,8,9\}} n_i + \frac{1}{3} F n_{10}
\end{eqnarray}

%\begin{eqnarray}
%    F_1 & = & F (n_1 + n_4 + n_5) + \frac{1}{3} F n_{10} \\
%    F_2 & = & F (n_2 + n_6 + n_7) + \frac{1}{3} F n_{10} \\
%    F_3 & = & F (n_3 + n_8 + n_9) + \frac{1}{3} F n_{10} \\
%\end{eqnarray}

Similarly we compute the torques applied through edge control points and
the central control points by:

\begin{gather}
    \begin{split}
        \theta_1 & = P_{1-4} \times (n_4 F) + P_{1-5} \times (n_5 F) + \\
        & + \frac{1}{3} (P_{1-4} + P_{1-5}) \times (n_{10} F)
    \end{split}
    \\
    \begin{split}
        \theta_2 & = P_{1-6} \times (n_6 F) + P_{1-7} \times (n_7 F) + \\
        & + \frac{1}{3} (P_{1-6} + P_{1-7}) \times (n_{10} F) \\
    \end{split}
    \\
    \begin{split}
        \theta_3 & = P_{1-8} \times (n_8 F) + P_{1-9} \times (n_9 F) + \\
        & + \frac{1}{3} (P_{1-8} + P_{1-9}) \times (n_{10} F) \\
    \end{split}
\end{gather}

%\begin{eqnarray}
%    \theta_1 & = & \sum_{i \in \{4,5\}} P_{1-i} \times (n_i F) \\
%    & + &\frac{1}{3} (P_{1-4} + P_{1-5}) \times (n_{10} F) \\
%    \theta_1 & = & \sum_{i \in \{6,7\}} P_{1-i} \times (n_i F) + \\
%    & + &\frac{1}{3} (P_{1-6} + P_{1-7}) \times (n_{10} F) \\
%    \theta_1 & = & \sum_{i \in \{8,9\}} P_{1-i} \times (n_i F) + \\
%    & + &\frac{1}{3} (P_{1-8} + P_{1-9}) \times (n_{10} F) \\
%\end{eqnarray}
%
%\begin{eqnarray}
%    \theta_1 & = & \sum_{i \in \{4,5\}} P_{1-i} \times (n_i F) +
%    \frac{1}{3} (P_{1-4} + P_{1-5}) \times (n_{10} F) \\
%    \theta_1 & = & \sum_{i \in \{6,7\}} P_{1-i} \times (n_i F) +
%    \frac{1}{3} (P_{1-6} + P_{1-7}) \times (n_{10} F) \\
%    \theta_1 & = & \sum_{i \in \{8,9\}} P_{1-i} \times (n_i F) +
%    \frac{1}{3} (P_{1-8} + P_{1-9}) \times (n_{10} F) \\
%\end{eqnarray}

@@

% We cannot show results for simple plane in bending because the solution
% seems to converge to wrong value (but it converges! :)). Thus showing
% results for plain membrane deformations doesn't make sense (it would be
% inconsistent).
  \CD{ - On this section,  we can maybe add some results from the CBM paper you did with Stephan especially  the figure 9 that illustrates the 3d contribution we list in the intro: i.e. Mapping of high resolution mesh on the curved surface of the shell.}

In next two subsections we present tests to validate the convergence of our
B\'ezier shell element. We use a combination of two elements that are very well
established in the area of mechanical modelling.
Our results are compared with the shell element
composed of Discrete Kirchoff Triangle (DKT) element \cite{Batoz1980} for bending plate and optimal ANDES element \cite{Felippa2003} for membrane.
The last subsection presents evaluation of the computational complexity of our solution.

\subsection{Roof Test}

To validate the results of the element we have performed a modified version
of a test known as Scordelis-Lo roof. It simulates a cylindrical roof under
self-weight. The geometry is defined by 80° cylindrical patch of length
$L=50$ and radius $r=25$ and thickness $t=0.25$ and is discretized into
mesh of NxN nodes. The physical parameters are $E = 4.32 \times 10^8$ and
$\nu = 0$. The roof is loaded with uniformly distributed load $q = 90$ per
unit area. The curved edges are clamped at both ends to avoid rigid body
movement. This the only difference from the original Scordelis-Lo roof test
where the edges are free to move in longitudinal direction. The Figure
\ref{fig-roof} shows the vertical displacement of midpoint on the free
edge.

@@

\subsection{Hemisphere Test}

Second performed test is a hemisphere with hole subjected to two opposing
forces at the base. The geometry is described by a hemisphere with radius
$r=10$ and thickness $t=0.04$ with 18° hole at the top and the geometry is
discretized into the grid of NxN nodes per quadrant. The physical
parameters are $E = 6.825 \times 10^7$ and $\nu = 0.3$. The hemisphere is
at it's base subjected to two opposing outwards forces and two opposing
inwards forces with magnitude $P = 4$. To avoid rigid body movement we have
constrained two opposite nodes at the top of the hemisphere. The radial
displacement for one point with applied load is shown in Figure
\ref{fig-hemisphere}.

@@

\subsection{Computation Speed}

In figure \ref{fig-fps} we present performance of the element in terms of
frames per second. All tests were performed on machine equipped Dual-Core
AMD Opteron Processor 2218 and 3 GB of RAM and conjugate gradient solver
has been to solve the system. Our implementation is without optimizations
and isn't parallel, which means only single CPU core was used for the
simulation. The values reported are for the raw physical simulation with
and without the frame fixing method described in subsection
\ref{sec-corot}. We also present results including the mapping of
high-resolution mesh onto the mechanical mesh for visualisation purposes or
collisions.

\Figure{\linewidth}{fps}
{Performance in frames per seconds for different number of elements. Shows
values for element with frame fixing (B) method described in subsection
\ref{sec-corot} and without it (A), and with high-resolution mesh of 2048
(C) or 3200 (D) triangles mapped onto the mechanical mesh.}

We can see that the frame fixing technique is really fast and we are able
to simulate up to 1000 elements and still maintain visual refresh rate of
25 FPS. Our implementation of mechanical mapping clearly needs improving
because it consumes most of the computation time, but it is still able to
provide good results for more than 600 elements.

\subsection{Application on Congenital Heart Disease Correction Planning}

@@

%%- - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
%
%% TODO: is this OK for anonymity?
%The presented element has been successfully implemented in SOFA framework
%\cite{Faure2012} for real-time surgical simulations. It has been evaluated
%on several applications.
%
%% TODO: !!! When ANONYMISING do something with the citation !!!
%One being a tool for pre-operative planning of corrective surgery of
%congenital heart diseases in infants \cite{Kislinskiy2012}. The cardiac
%surgery is a complicated process where several arterial vessels have to be
%cut, patched and reconnected. The element has been used for to model
%deformations of blood vessels after suturing and/or applying a artificial
%patch. Figure \ref{fig-chd} presents a screenshots of the application
%prototype.

>>>>

% Partial derivative
\newcommand{\deriv}[2]{\frac{\partial #1}{\partial #2}}
% Matrix
\newcommand{\mat}[1]{\mathbf{#1}}

@@

\section{Element Kinematics}\label{sec-kinematics} % {{{

@@

\begin{equation}\label{eq-bez-delta}
    \begin{split}
        \delta T = &   (n_1 + n_4 + n_5 + \frac{1}{3} n_{10})  \delta \mathbf{x}_1 + \\
                   & + (n_2 + n_6 + n_7 + \frac{1}{3} n_{10})  \delta \mathbf{x}_2 + \\
                   & + (n_3 + n_8 + n_9 + \frac{1}{3} n_{10})  \delta \mathbf{x}_3 + \\
                   & + \sum_{j=4}^9 n_j P_{i-j} \times \delta \mathbf{\theta}_i + \\
                   & + \frac{n_{10}}{3} \sum_{i=1}^3 P_{i-10} \times \delta \mathbf{\theta}_i 
    \end{split}
\end{equation}

@@

The term $\deriv{T^k}{\xi_i}$ can be derived from the expression of $T(\xi_1,\xi_2, \xi_3)$ in equation (\ref{global-transform}) and 
$\deriv{\xi_i}{l}$ from equation (\ref{shape-vector}). 
The in-plane displacement gradient can be obtained with tensor $F$:
%
\begin{equation}
\deriv{u^k}{l} = F_{kl} - I_{kl} 
\end{equation}
%
where $\mat{I}$ is the identity matrix ($I_{kl}  = \{1$ if $l=k$ ;  $0$ if $l \neq k \} $). 
The displacement gradient can be linked to the degrees of freedom of the shell using the expression of the displacement of a point $\delta T$ (see equation  (\ref{eq-bez-delta})) on the shell surface:
\begin{equation}
    \nabla u = \nabla(\delta T)
\end{equation}

@@

In order to compute FEM-based stiffness for the shell element, we need to compute the jacobian matrix between strains and nodes' displacements.
As we have separated the formulations for the in-plane and the out of plane displacements, it will result with the computation of two jacobian matrices (of size $3\times9$).
\begin{equation}
    \delta 
    \begin{bmatrix}
        \epsilon_{xx} \\ \epsilon_{yy}  \\ 2\epsilon_{yx}
    \end{bmatrix}
    = \begin{bmatrix} \mat{J}_m \end{bmatrix}
    \begin{bmatrix}
        \delta \mat{x}_1^x \\ \delta \mat{x}_1^y \\ \delta \theta_1^z \\ \vdots
    \end{bmatrix}
\end{equation}
where $\left[ \mat{J}_m \right]$ maps the membrane in-plane displacements and 
%
\begin{equation}
    \delta
    \begin{bmatrix}
        \epsilon_{xx} \\ \epsilon_{yy}  \\ 2\epsilon_{yx}
    \end{bmatrix}
    = \begin{bmatrix} \mat{J}_b \end{bmatrix}
    \begin{bmatrix}
        \delta \mat{x}_1^z \\ \delta \theta_1^x \\ \delta \theta_1^y \\ \vdots
    \end{bmatrix}
\end{equation}
where $\left[ \mat{J}_b \right]$ maps the bending displacements.

To compute the values in matrices $\left[ \mat{J}_m \right]$ and $\left[ \mat{J}_b \right]$,  we can reuse equation (\ref{eq-bez-delta}) and apply the spatial derivatives.
To illustrate, let's take one value of the matrix $\left[ \mat{J}_m \right]$  (first line, fourth column), we first take the kinematics between $\delta T_x$ and the concerned degree of freedom $ \delta \mathbf{x}_2^x$ from equation (\ref{eq-bez-delta}):
\begin{equation}
    \frac{ \delta T_x }{ \delta \mathbf{x}_2^x } = (n_2 + n_6 + n_7 + \frac{n_{10}}{3})
\end{equation}

Then, we apply the spatial derivatives to compute the variation of $\epsilon_{xx}$ due to a variation of $\mathbf{x}_{2x}$:
\begin{equation}
    \delta \epsilon_{xx} += 
    \underbrace{\left(
    \sum_{i=1}^{3} (\deriv{n_2}{\xi_i} + \deriv{n_6}{\xi_i} + \deriv{n_7}{\xi_i} + \frac{1}{3}\deriv{n_{10}}{\xi_i})
    \deriv{\xi_i}{x} \right) }_{\begin{bmatrix} \mat{J}_m \end{bmatrix}_{14}}  
 \delta \mathbf{x}_2^x 
\end{equation}

@@

\begin{align}
    \label{eq-Km}
    \mat{K}_m & = \iiint_V \mat{J}_m^T \mat{M} \mat{J}_m \, \mathrm{d} V \\
    \label{eq-Kb}
    \mat{K}_b & = \iiint_V \mat{J}_b^T \mat{M} \mat{J}_b \, \mathrm{d} V
\end{align}

\noindent
where $\mat{M}$ is the material matrix. To keep the system simple we use linear
Hooke's law for isotropic materials.

Because the deflection field for the element is non-linear \eqref{eq-bez-delta}
in position the integrals \eqref{eq-Km} and \eqref{eq-Kb} have to be computed using
numerical integration. In our implementation we employ 6-point Gaussian
quadrature for integration over triangle area.
\CD{TODO: figure + gives the formula of the quadrature...}
% TODO: add ref
% TODO: add abscissas/weights?

@@

It is known that the Cauchy's strain tensor \eqref{eq-in-plane-strain} is not
rotation invariant \cite{Hauth2004,Muller2004} and produces ghost forces for rigid
rotations. The alternative approach is to use non-linear Green's strain tensor but the non-linearity introduced into the system is costly and introduces instabilities.
To keep the system linear and to deal
with rigid body rotations we compute the displacements and forces in
corotational frame. At each step of the simulation we compute the local
frame for every element, the resulting internal force of deformed element
is then computed relative to this frame:
%
\begin{equation}
    F_e = \mat{R}_e^T (\mat{K}_e u) = \mat{R}_e^T \left(\mat{K}_e(\mat{R}_e x - \mat{R}_e^0 x^0)\right)
\end{equation}
%
where $\mat{R}$ and $\mat{R}_0$ are rotation matrices of the frame for current and rest
mesh respectively.

The corotational frame is computed followingly:

\begin{itemize}
    \item One axis is alligned to one edge of the element:\\
    $\mathbf{e}_1 = \mathbf{x}_2 - \mathbf{x}_1$
    \item A normal to the plane of the element is computed:\\
    $\mathbf{e}_3 = \mathbf{e}_1 \times (\mathbf{x}_3 - \mathbf{x}_1)$
    \item Second axis in the plane of the element is computed:\\
    $\mathbf{e}_2 = \mathbf{e}_1 \times \mathbf{e}_3$
\end{itemize}

Correct choice of the corotational frame is a tricky
problem \cite{Felippa2005} and incorrect frame for the membrane element can
have adverse effects on the simulation.
This choice of the frame where we align one axis with one of the edges
of the triangle is not a perfect one and we have experienced
nonnegligible difference in deformation of equilateral triangle depending
on whether an axis was aligned to the edge favored by the deformation or to
one of the two other edges.

A polar decomposition on the deformation gradient $F = \nabla u
+ \mat{I}$ can be used to extract a rotation component from the deformation of
the element \cite{Ciarlet1994,Hauth2004}. Because we are only fixing the membrane
element we can perform the polar decomposition in 2D with the following
formula:
%
\begin{equation}
    \mat{R} = \mat{F} + \mathrm{sgn}(\mathrm{det}(\mat{F})) \begin{pmatrix}
        F_{22} & - F_{21} \\
        F_{12} & F_{11}
    \end{pmatrix}
\end{equation}
%
and subsequently normalizing the columns. We then rotate the corotational
frame in opposite direction. Even single step of this process improves the
final frame but more iterations of the polar decompositions are usually
necessary (up to 20) which severely degrades the performance. Polar
decomposition tends to "overshoot" the ideal rotation angle and each
subsequent step performs a rotation in the direction opposite to the rotation in previous step. By properly
scaling the rotation angle we are able to minimize the amount of needed
iterations down to 5. The best value for the scaling factor depends on the
simulated problem, but we have experimentally localized the best value to
be somewhere between 0.6 and 0.65. We choose to use the value 0.61.

We employ two stop conditions for the iteration process:
(i) the change in rotation angle is less than $10^{-6}$, and
(ii) the maximum number of iterations is set to 5.

If the first condition is not met in 5 iterations it can mean one of three
things we still have very good approximation of the ideal frame. In some
situations the process is not able to reach the ideal frame quickly because
either the convergence for the element is too slow or the process is diverging.
In either case performing more iterations is unnecessary.

@@

\section{Application of High Resolution Mesh} % {{{

Because of the bending property of the shells relatively few elements are
necessary to simulate curved surface. To enrich the visual experience from
the simulated object it is desirable to use more triangles in the areas
with high curvature during rendering. Besides the visual accuracy the high
resolution mesh correctly modeling the curved areas is needed for
interaction and correct simulation of collisions and constraints. To handle
these we also need to know how to map the velocities from the mechanical
mesh onto the high resolution mesh and how to map forces acting on the high
resolution mesh back onto the mechanical mesh.

\subsection{Surface Mesh}

To project vertices onto the surface of shells 
we use the fact that the geometry of is based on the formulation
of cubic B\'ezier triangle \eqref{eq-cubicbez}. For every vertex of the high
resolution mesh we first find the corresponding triangle on the mechanical
mesh that is closest to the vertex and assign barycentric coordinates on
the triangle to this vertex. After every step of the simulation points of the
control mesh are updated based on the kinematic link described in
section \ref{sec-kinematics} and the high
resolution mesh is updated using the assigned barycentric coordinates and
the function of the surface \eqref{eq-cubicbez}.

\subsection{Nodal Velocities of the Surface Mesh}

For contact modeling it is often necessary to know the velocity of each point.
The values are directly available to us for the nodes of the mechanical mesh
but they can also be computed for any point on the surface.
Differentiating \eqref{eq-cubicbez2} by time we get:
%
\begin{equation}
    \dot{T} = \sum_{i=1}^{10} n_i \dot{P_i} = \sum_{i=1}^{10} n_i V_i
\end{equation}

Which means we use the same expression to interpolate the velocity only by
substituting the velocities at control points. For the corner nodes we
already have the velocities $V_i$. For the internal nodes we need to compute
them. We again make use of the kinematic link between nodes and solve the
problem of computing the velocity of a point attached to the moving rigid body.
For the edge points we have:
%
\begin{eqnarray}
    V_j = V_i + \omega_i \times P_{i-j}
\end{eqnarray}
%
where $i$ is the index of the corner node the edge node is attached to and
$\omega_i$ is the angular velocity at the node. For the central node
the formula is based on \eqref{eq-central2}:
%
\begin{eqnarray}
    V_i = \frac{1}{3} \sum_{i=1}^3 \left( V_i + \omega_i \times P_{i-10} \right)
\end{eqnarray}

\subsection{Projection of Normal Forces}

For physical interaction with the object one also requiers that the normal
forces applied on the object's surface are trasmitted back as forces
and torques on the vertices of the mechanical mesh.
We compute the influence of force $F$ on the corners through all the
associated control points by the formula:
%
\begin{eqnarray}
    F_1 & = & F \sum_{i \in \{1,4,5\}} n_i + \frac{n_{10}}{3} F \\
    F_2 & = & F \sum_{i \in \{2,6,7\}} n_i + \frac{n_{10}}{3} F \\
    F_3 & = & F \sum_{i \in \{3,8,9\}} n_i + \frac{n_{10}}{3} F 
\end{eqnarray}

Similarly we can compute the torques applied through edge control points and
the central control points:
%
\begin{gather}
    \begin{split}
        \theta_1 = & P_{1-4} \times (n_4 F) + P_{1-5} \times (n_5 F) + \\
        & + \frac{1}{3} P_{1-10} \times (n_{10} F)
    \end{split}
    \\
    \begin{split}
        \theta_2 = & P_{2-6} \times (n_6 F) + P_{2-7} \times (n_7 F) + \\
        & + \frac{1}{3} P_{2-10} \times (n_{10} F) \\
    \end{split}
    \\
    \begin{split}
        \theta_3 = & P_{3-8} \times (n_8 F) + P_{3-9} \times (n_9 F) + \\
        & + \frac{1}{3} P_{3-10} \times (n_{10} F) \\
    \end{split}
\end{gather}

@@

% We cannot show results for simple plane in bending because the solution
% seems to converge to wrong value (but it converges! :)). Thus showing
% results for plain membrane deformations doesn't make sense (it would be
% inconsistent).

% TODO:
% \CD{ - On this section,  we can maybe add some results from the CBM paper you did with Stephan especially  the figure 9 that illustrates the 3d contribution we list in the intro: i.e. Mapping of high resolution mesh on the curved surface of the shell.}
% Not enough space?

In subsections \ref{sec-roof} and \ref{sec-hemisphere} we present tests to
validate the convergence of our B\'ezier shell element. As a reference we use a
combination of two elements that are very well established in the area of
mechanical modelling. Our results are compared with the shell element composed
of Discrete Kirchoff Triangle (DKT) element \cite{Batoz1980} for bending plate
and optimal ANDES element \cite{Felippa2003} for membrane. For comparison the
results for the model of O. Comas are also presented. Subsection \ref{sec-fps}
presents evaluation of the computational complexity of our solution. Last
subsections are dedicated to presentation of application of our model.

\subsection{Roof Test}\label{sec-roof}

To validate the results of the element we have performed a modified version
of a test known as Scordelis-Lo roof. It simulates a cylindrical roof under
self-weight. The geometry is defined by 80° cylindrical patch of length
$L=50$, radius $r=25$ and thickness $t=0.25$ and is discretized into
mesh of NxN vertices. The physical parameters are $E = 4.32 \times 10^8$ and
$\nu = 0$. The roof is loaded with uniformly distributed load $q = 90$ per
unit area. The curved edges are clamped at both ends to avoid rigid body
movement. This is the only difference from the original Scordelis-Lo roof test
where the edges are free to move in longitudinal direction. The Figure
\ref{fig-roof} shows the vertical displacement of midpoint on the free
edge.

@@

\subsection{Hemisphere Test}\label{sec-hemisphere}

Second performed test is a hemisphere with hole subjected to two opposing
forces at the base. The geometry is described by a hemisphere with radius
$r=10$ and thickness $t=0.04$ with 18° hole at the top and the geometry is
discretized into the grid of NxN vertices per quadrant. The physical
parameters are $E = 6.825 \times 10^7$ and $\nu = 0.3$. The hemisphere is
at it's base subjected to two opposing outwards forces and two opposing
inwards forces with magnitude $P = 4$. To avoid rigid body movement we have
constrained two opposite nodes at the top of the hemisphere. The radial
displacement for one point with applied load is shown in Figure
\ref{fig-hemisphere}.

@@

\subsection{Computation Speed}\label{sec-fps} % {{{

In figure \ref{fig-fps} we present performance of the element in terms of
frames per second. All tests were performed on machine equipped with Dual-Core
AMD Opteron Processor 2218 and 3~GB of RAM. We used a conjugate gradient solver
to solve the system. Our implementation is without optimizations
and isn't parallel, which means only single CPU core was used for the
simulation. The values reported are for the raw physical simulation with (b)
and without (a) the frame fixing method described in subsection
\ref{sec-corot}. We also present results including the mapping of
high-resolution mesh onto the mechanical mesh for visualisation purposes or
collisions.

\Figure{\linewidth}{fps}
{Performance in frames per seconds for different number of elements. Shows
values for element with frame fixing (B) method described in subsection
\ref{sec-corot} and without it (A), and with mesh of 3\,200
(C) or 10\,082 (D) triangles mapped onto the mechanical mesh.}

We can see that the frame fixing technique is really fast and we are able
to simulate up to 1\,000 elements and still maintain visual refresh rate of
25 FPS. While our implementation was only single-threaded and the mechanical mapping
wasn't off-loaded to another CPU core we were still able to provide good
results for more than 600 elements mesh with 10k triangles.

\subsection{Application on Congenital Heart Disease Correction Planning} % {{{

@@

%%- - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
%
%% TODO: is this OK for anonymity?
%The presented element has been successfully implemented in SOFA framework
%\cite{Faure2012} for real-time surgical simulations. It has been evaluated
%on several applications.
%
